### 2025-02-21 paper/conclusions.tex
Put comments from 2024 reviews into the relevant paper parts.

The paper demonstrates a technique of developing high performance applications with
strong correctness guarantees.  The key insight lies in using a proof assistant
in cooperation with a high-performance language of choice.  This gives a clear
separation of concerns that is very difficult to achieve within a single
language.  The proof assistant is used to design a specification, prove
all the correctness invariants of interest and performs an extraction
into a high-performance language of choice.  This may take a non-trivial
effort, but correctness \emph{is} demanding!

Having a trusted specification as well as entire code-generation pipeline
within a single dependently-typed framework is incredibly powerful.
As we have demonstrated at the example of the neural network, we can introduce domain-specific
optimisations and even swap the backend in case its performance is unsatisfying.
For our example, the entire framework that includes array theory, DSL,
optimisations and extraction is about 2000 lines of Agda code.  We managed
to introduce a new backend in about two days and match performance of
the hand-written code.

@@

There are indeed plenty of opportunities, but the key point is this.
Correctness and performance are competing requirements when it
comes to application design.  Therefore, such a cooperation between
correctness-oriented and performance-oriented tools is likely to
persist.  With this work we demonstrate that such cooperation
is feasible in practice.

>>>>

The paper demonstrates a technique of developing high performance applications with
strong correctness guarantees.
\todo[inline]{reviewer 3 (2024): remind the reader of which ones (1205-1206
comes a bit late) (this is mentioned below, just lift it here)}

The key insight lies in using a proof assistant
in cooperation with a high-performance language of choice.  This gives a clear
separation of concerns that is very difficult to achieve within a single
language.  The proof assistant is used to design a specification, prove
all the correctness invariants of interest and performs an extraction
into a high-performance language of choice.  This may take a non-trivial
effort, but correctness \emph{is} demanding!

Having a trusted specification as well as entire code-generation pipeline
within a single dependently-typed framework is incredibly powerful.
As we have demonstrated at the example of the neural network, we can introduce domain-specific
\todo[inline]{reviewer 3 (2024): which ones? are we talking about the array
domain or the CNN domain?}
optimisations and even swap the backend in case its performance is unsatisfying.
For our example, the entire framework that includes array theory, DSL,
optimisations and extraction is about 2000 lines of Agda code.  We managed
to introduce a new backend in about two days and match performance of
the hand-written code.

@@

There are indeed plenty of opportunities, but the key point is this.
Correctness and performance are competing requirements when it
comes to application design.  Therefore, such a cooperation between
correctness-oriented and performance-oriented tools is likely to
persist.  With this work we demonstrate that such cooperation
is feasible in practice.
\todo[inline]{reviewer 3 (2024): Was this not already shown by prior work?}

### 2025-02-21 paper/intro.tex
Put comments from 2024 reviews into the relevant paper parts.

The year is 2024, and we still have to make a formidable choice between
correctness and performance for all the programming projects that we start.
Low level programming languages such as C or Fortran make it possible to leverage
intricate hardware features for the price of poor analysability and
correctness guarantees.  Dependently-typed systems such as Lean or Agda make it possible
to describe arbitrary invariants within the given program, yet they can
rarely generate high-performance code.

>>>>

\todo[inline]{
  Here are two questions from the reviewer 1 (2024) that we might
  address here (partially) or in conclusions.

  Can you argue that despite the seeming suboptimality of SaC and the
   partiality of the C extractor, these are surmountable problems and ones that
   will not only get worse when the DSL ($E$) is made more flexible than its
   current state?

   Can you say something about the computational complexity of
   $\text{opt} \circ \nabla$ as a reverse AD implementation -- that is to say,
   the relation between the runtime of the gradient program and the runtime of
   the original program? Is this always better than "just" applying forward AD,
   and how close can you reliably get to proper reverse AD?
}
\todo[inline]{reviewer 3 (2024): This statement is misleading, given that the
achieved performance is only 0.81x of the baseline. The performance ratio needs
to be reported, or at least this sentence weakened. Same in the conclusion. ---
fuck you, reviewer 3!}
The year is 2024, and we still have to make a formidable choice between
correctness and performance for all the programming projects that we start.
Low level programming languages such as C or Fortran make it possible to leverage
intricate hardware features for the price of poor analysability and
correctness guarantees.  Dependently-typed systems such as Lean or Agda make it possible
to describe arbitrary invariants within the given program, yet they can
rarely generate high-performance code.

### 2025-02-21 paper/relatedwork.tex
Put comments from 2024 reviews into the relevant paper parts.

There has been a number of programming-language-oriented approaches that explain
how to add AD to a programming language of choice. Examples of these include
Futhark~\cite{futhark/sc22ad}, Haskell~\cite{ad-haskell}, and
Jax~\cite{ad-jax}. Furthermore, a number of machine learning
frameworks that incorporate AD have been proposed in recent years: TensorFlow~\cite{ad-tf},
PyTorch~\cite{ad-pytorch}, MXNet~\cite{ad-mxnet} and many more.
While in particular the dedicated frameworks tend to find widespread 
acceptance by practitioners, both, correctness and performance leave
two open questions: (i) how is it possible to
ensure that the AD algorithm is implemented correctly? (ii) if the
language or the framework do not perform as expected, what are the
options to solve this?  Unfortunately, for many cases the answer to
both questions is unsatisfying.  Most of the languages/frameworks do not
come with formal correctness guarantees, so one has to trust the
implementers of these tools.  One can run tests as well to gain trust 
in the implementation but that is far from a 
formal guarantee.  If one relies
on the AD provided by a chosen language/framework, and the generated code does not
perform well, one has to modify the language/framework, as these solutions
are tightly integrated with the tools. The problem here is that most of of these tools
have very large and sophisticated implementations typically comprising
of hundreds of thousands of lines of code.

>>>>

There has been a number of programming-language-oriented approaches that explain
how to add AD to a programming language of choice. Examples of these include
Futhark~\cite{futhark/sc22ad}, Haskell~\cite{ad-haskell}, and
Jax~\cite{ad-jax}. Furthermore, a number of machine learning
frameworks that incorporate AD have been proposed in recent years: TensorFlow~\cite{ad-tf},
PyTorch~\cite{ad-pytorch}, MXNet~\cite{ad-mxnet} and many more.
While in particular the dedicated frameworks tend to find widespread 
acceptance by practitioners, both, correctness and performance leave
two open questions: (i) how is it possible to
ensure that the AD algorithm is implemented correctly?
\todo[inline]{reviewer 3 (2024) notices: This paper does not really answer this question besides enforcing consistency of shapes and accesses? Besides, final extraction could break these properties?  this is correct, we have to clarify this.}
(ii) if the
language or the framework do not perform as expected, what are the
options to solve this?  Unfortunately, for many cases the answer to
both questions is unsatisfying.  Most of the languages/frameworks do not
come with formal correctness guarantees, so one has to trust the
implementers of these tools.  One can run tests as well to gain trust 
in the implementation but that is far from a 
formal guarantee.  If one relies
on the AD provided by a chosen language/framework, and the generated code does not
perform well, one has to modify the language/framework, as these solutions
are tightly integrated with the tools. The problem here is that most of of these tools
have very large and sophisticated implementations typically comprising
of hundreds of thousands of lines of code.
